Reanalysis Step 4: Outcome Analysis with Three-Way Interaction

### 2025-08-19 reanalysis-2-multiple_imputation.qmd
better diagnostics and cleaned up code

# Set parameters based on mode
if (analysis_mode == "test") {
  n_imputations <- 2
  n_iterations <- 2
  message("Running in TEST mode: ", n_imputations, " imputations, ", n_iterations, " iterations")
} else if (analysis_mode == "final") {
  n_imputations <- 10
  n_iterations <- 20
  message("Running in FINAL mode: ", n_imputations, " imputations, ", n_iterations, " iterations")
} else {
  stop("Invalid analysis_mode. Must be 'test' or 'final'")
}
```

@@

  message("Proportion of DR severity across imputations:")
  message("No_DR: ", round(mean(dr_props[1,]), 3), " (SD: ", round(sd(dr_props[1,]), 3), ")")
  message("NPDR: ", round(mean(dr_props[2,]), 3), " (SD: ", round(sd(dr_props[2,]), 3), ")")
  message("PDR: ", round(mean(dr_props[3,]), 3), " (SD: ", round(sd(dr_props[3,]), 3), ")")

  # Visual check if stripplot works with factors
  tryCatch({
    stripplot(mice_results, person_dr ~ .imp, pch = 20, cex = 1.2) |> print()
  }, error = function(e) {
    message("Note: Stripplot not available for categorical person_dr")
  })
}

@@

  message("Proportion of cataract surgery across imputations:")
  if(is.matrix(cat_props)) {
    for(i in 1:nrow(cat_props)) {
      message("Level ", rownames(cat_props)[i], ": ", 
              round(mean(cat_props[i,]), 3), " (SD: ", round(sd(cat_props[i,]), 3), ")")
    }
  } else {
    message("Prop = ", round(mean(cat_props), 3), " (SD: ", round(sd(cat_props), 3), ")")
  }
}
```

@@

# Variables to add back (those not in imputation model or created by transformations)
vars_in_imputed <- names(imputed_datasets[[1]])
vars_to_add <- setdiff(names(df_original_full), vars_in_imputed)

if(length(vars_to_add) > 0) {
  message("Adding back ", length(vars_to_add), " variables not in imputation model...")

  # Add these variables back to each imputed dataset
  imputed_datasets <- lapply(imputed_datasets, function(imp_data) {
    imp_data <- imp_data |>
      left_join(
        df_original_full |> 
          dplyr::select(e_mrn_deidentified, dplyr::all_of(vars_to_add)),
        by = "e_mrn_deidentified",
        relationship = "many-to-many"  # Expected: multiple eyes per patient
      )

>>>>

# Set parameters based on mode
if (analysis_mode == "test") {
  n_imputations <- 2
  n_iterations <- 2
} else if (analysis_mode == "final") {
  n_imputations <- 10
  n_iterations <- 20
} else {
  stop("Invalid analysis_mode. Must be 'test' or 'final'")
}
```

**Running in `r toupper(analysis_mode)` mode:**  
- **Imputations:** `r n_imputations`  
- **Iterations:** `r n_iterations`

@@

  # Create summary table
  dr_summary <- data.frame(
    Category = c("No_DR", "NPDR", "PDR"),
    Mean_Proportion = c(round(mean(dr_props[1,]), 3), 
                       round(mean(dr_props[2,]), 3),
                       round(mean(dr_props[3,]), 3)),
    SD = c(round(sd(dr_props[1,]), 3),
           round(sd(dr_props[2,]), 3), 
           round(sd(dr_props[3,]), 3))
  )
  kable(dr_summary, caption = "DR severity proportions across imputations")

  # Visual check if stripplot works with factors
  tryCatch({
    stripplot(mice_results, person_dr ~ .imp, pch = 20, cex = 1.2) |> print()
  }, error = function(e) {
    # Note: Stripplot not available for categorical person_dr
  })
}

@@

  # Create summary table for cataract surgery proportions
  if(is.matrix(cat_props)) {
    cat_summary <- data.frame(
      Level = rownames(cat_props),
      Mean_Proportion = apply(cat_props, 1, mean) |> round(3),
      SD = apply(cat_props, 1, sd) |> round(3)
    )
  } else {
    cat_summary <- data.frame(
      Level = "Cataract Surgery",
      Mean_Proportion = round(mean(cat_props), 3),
      SD = round(sd(cat_props), 3)
    )
  }
  kable(cat_summary, caption = "Cataract surgery proportions across imputations")
}
```

@@

# Variables to add back (those not in imputation model or created by transformations)
vars_in_imputed <- names(imputed_datasets[[1]])
vars_to_add <- setdiff(names(df_original_full), vars_in_imputed)
```

### Adding Back Non-Imputed Variables

**Variables to add back:** `r if(length(vars_to_add) > 0) length(vars_to_add) else 0` variables not in imputation model

```{r add-back-vars}
if(length(vars_to_add) > 0) {
  # Add these variables back to each imputed dataset
  imputed_datasets <- lapply(imputed_datasets, function(imp_data) {
    imp_data <- imp_data |>
      left_join(
        df_original_full |> 
          dplyr::select(e_mrn_deidentified, dplyr::all_of(vars_to_add)),
        by = "e_mrn_deidentified",
        relationship = "many-to-many"  # Expected: multiple eyes per patient
      )

### 2025-08-19 reanalysis-3-propensity_scores.qmd
better diagnostics and cleaned up code

# Set parameters based on mode
if (analysis_mode == "test") {
  max_imputations_for_diagnostics <- 2  # Show diagnostics for 2 imputations in test
  n_trees <- 1000  # Fewer trees for testing
} else {
  max_imputations_for_diagnostics <- 5  # Show diagnostics for up to 5 imputations in final
  n_trees <- 3000  # More trees for final analysis
}

@@

**Analysis mode:** `r analysis_mode`  
**Will show diagnostics for up to** `r max_imputations_for_diagnostics` **imputations**  
**Using** `r n_trees` **trees for GBM**

@@

```{r load-data}
imputed_datasets <- readRDS(file.path(reanalysis_data_dir, "imputed_datasets.rds"))
n_imputations <- length(imputed_datasets)

@@

**Loaded** `r n_imputations` **imputed datasets**

@@

n_to_show <- min(max_imputations_for_diagnostics, n_imputations)
imps_to_show <- 1:n_to_show

# Generate diagnostic plots for selected imputations
for (i in imps_to_show) {

@@

  # Print section header
  cat("\n### Imputation", i, "\n\n")

  # Plot 1: Iteration convergence
  cat("#### Convergence\n")
  plot(twang_model, plots = 1)

  # Plot 2: Boxplots of weights
  cat("\n#### Weight Distribution\n")
  plot(twang_model, plots = 2)

  # Plot 3: Balance reduction (standardized effect sizes)
  cat("\n#### Covariate Balance\n")
  plot(twang_model, plots = 3)

  # Plot 6: Proportion of weights histogram
  cat("\n#### Weight Proportions\n")
  plot(twang_model, plots = 6)

  # Display balance summary table
  cat("\n#### Balance Summary Table\n")
  print(summary(twang_model))
}
```

>>>>

# Set parameters based on mode
if (analysis_mode == "test") {
  max_imputations_to_process <- 2  # Process only 2 imputations in test mode
  max_imputations_for_diagnostics <- 2  # Show diagnostics for 2 imputations in test
  n_trees <- 1000  # Fewer trees for testing
} else {
  max_imputations_to_process <- NULL  # Process all available imputations
  max_imputations_for_diagnostics <- 5  # Show diagnostics for up to 5 imputations in final
  n_trees <- 3000  # More trees for final analysis
}

@@

**Analysis mode:** `r analysis_mode`  
**Imputations to process:** `r if(!is.null(max_imputations_to_process)) max_imputations_to_process else "all available"`  
**Diagnostics shown for:** `r max_imputations_for_diagnostics` **imputations**  
**GBM trees:** `r n_trees`

@@

```{r load-data}
imputed_datasets <- readRDS(file.path(reanalysis_data_dir, "imputed_datasets.rds"))
n_available <- length(imputed_datasets)

# Determine how many imputations to actually process
if (!is.null(max_imputations_to_process)) {
  n_imputations <- min(max_imputations_to_process, n_available)

  if (n_imputations < n_available) {
    imputed_datasets <- imputed_datasets[1:n_imputations]
  }
} else {
  n_imputations <- n_available
}

@@

**Available imputed datasets:** `r n_available`  
**Processing:** `r n_imputations` **datasets** `r if(analysis_mode == "test") "(test mode limit)" else "(all available)"`

@@

n_to_show <- min(max_imputations_for_diagnostics, n_imputations)
imps_to_show <- 1:n_to_show
```

```{r diagnostics-loop}
# Generate diagnostic plots for selected imputations
for (i in imps_to_show) {
  cat("\n### Imputation", i, "\n\n")

@@

  cat("#### Convergence\n")
  plot(twang_model, plots = 1)

  cat("\n#### Weight Distribution\n")
  plot(twang_model, plots = 2)

  cat("\n#### Covariate Balance\n")
  plot(twang_model, plots = 3)

  cat("\n#### Weight Proportions\n")
  plot(twang_model, plots = 6)

  cat("\n#### Balance Summary Table\n")
  print(summary(twang_model))
}
```

### 2025-08-19 reanalysis-4-outcome_analysis.qmd
refactor to be more quarto friendly
better diagnostics and cleaned up code

```{r setup}
#| include: false
# Setup analysis environment
source("dependencies.R")
setup_analysis(seed = 2025)

# Additional packages
library(emmeans)
library(margins)
library(survey)
library(ggplot2)
library(gridExtra)
library(knitr)
library(tidyr)    # For pivot_wider
library(dplyr)    # For data manipulation
library(mitools)  # For proper MI pooling with survey weights
library(mice)     # For pooling utilities

# ANALYSIS MODE: Use same as imputation/PS
analysis_mode <- "test"  # Should match imputation and PS modes

message("Analysis mode: ", analysis_mode)
```

@@

message("Loaded ", n_imputations, " imputed datasets with IPTW weights")

# Summary of first dataset for verification
first_data <- imputed_datasets[[1]]
message("First dataset summary:")
message("  - N observations: ", nrow(first_data))
message("  - N unique patients: ", length(unique(first_data$e_mrn_deidentified)))
message("  - Mean IPTW weight: ", round(mean(first_data$iptw_weight), 3))
message("  - Weight range: [", round(min(first_data$iptw_weight), 3), ", ", 
        round(max(first_data$iptw_weight), 3), "]")
```

@@

# Check distributions in first dataset
message("Variable distributions in first imputed dataset:")
message("DR Severity:")
print(table(imputed_datasets[[1]]$dr_severity))
message("Treatment Status:")
print(table(imputed_datasets[[1]]$any_treatment))
message("Lapse Status:")
print(table(imputed_datasets[[1]]$ever_lapse_binary))

# Cross-tabulation of the three factors
message("Three-way cross-tabulation:")
with(imputed_datasets[[1]], 
     ftable(dr_severity, any_treatment, ever_lapse_binary))
```

@@

message("Model formula:")
print(interaction_formula)

# Count interaction terms
message("Interaction structure:")
message("  - Main effects: 3 (lapse, DR severity [2 df], treatment)")
message("  - Two-way interactions: 3 (lapse×DR [2 df], lapse×treatment [1 df], DR×treatment [2 df])")
message("  - Three-way interaction: 1 (lapse×DR×treatment [2 df])")
message("  - Total interaction df: 7")
```

@@

message("Fitting weighted models to each imputed dataset...")

for (i in 1:n_imputations) {
  message("Imputation ", i, " of ", n_imputations, ":")

>>>>

```{r setup}
#| include: false
source("dependencies.R")
setup_analysis(seed = 2025)

# Additional packages
library(emmeans)
library(margins)
library(survey)
library(ggplot2)
library(gridExtra)
library(knitr)
library(tidyr)
library(dplyr)
library(mitools)
library(mice)

# ANALYSIS MODE: Use same as imputation/PS
analysis_mode <- "test"  # Should match imputation and PS modes
```

**Analysis mode:** `r analysis_mode`

@@

# Summary of first dataset for verification
first_data <- imputed_datasets[[1]]

dataset_summary <- data.frame(
  Metric = c("Imputed datasets loaded", "N observations", "N unique patients", 
             "Mean IPTW weight", "Weight range"),
  Value = c(n_imputations,
            nrow(first_data),
            length(unique(first_data$e_mrn_deidentified)),
            round(mean(first_data$iptw_weight), 3),
            paste0("[", round(min(first_data$iptw_weight), 3), ", ",
                   round(max(first_data$iptw_weight), 3), "]"))
)

kable(dataset_summary, caption = "Dataset Summary")
```

@@

```

### Variable Distributions in First Imputed Dataset

**DR Severity:**
```{r dr-severity-dist}
table(imputed_datasets[[1]]$dr_severity)
```

**Treatment Status:**
```{r treatment-dist}
table(imputed_datasets[[1]]$any_treatment)
```

**Lapse Status:**
```{r lapse-dist}
table(imputed_datasets[[1]]$ever_lapse_binary)
```

**Three-way Cross-tabulation:**
```{r crosstab}
with(imputed_datasets[[1]], 
     ftable(dr_severity, any_treatment, ever_lapse_binary))
```

@@

```

### Model Specification

**Formula:**
```{r show-formula}
print(interaction_formula)
```

**Interaction Structure:**
- Main effects: 3 (lapse, DR severity [2 df], treatment)
- Two-way interactions: 3 (lapse×DR [2 df], lapse×treatment [1 df], DR×treatment [2 df])
- Three-way interaction: 1 (lapse×DR×treatment [2 df])
- Total interaction df: 7

@@

# Track model fitting progress
model_progress <- data.frame(
  Imputation = integer(),
  Converged = logical(),
  ThreeWay_Terms = integer()
)

for (i in 1:n_imputations) {
